---
title: "Koreatown Homelessness and Rental Cost Analysis"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

# Research Questions:

1.  How are increases in rental housing costs associated with changes in
    different categories of unsheltered homelessness (vehicle,
    tent/encampment, street) in Koreatown before, during, and after
    COVID-19?
2.  Did rising rental costs affect rates of single-adult vs. family
    homelessness differently in Koreatown across the same time periods?

## Koreatown ZIP Codes: 90004, 90005, 90006, 90010

```{r setup, message=FALSE, warning=FALSE}
# Import necessary libraries
library(tidyverse)
library(httr)
library(jsonlite)
library(scales)
library(gridExtra)
library(reshape2)
library(corrplot)
library(httr)

# Set plotting theme
theme_set(theme_minimal())

print("Libraries imported successfully!")
```

## 1. Load and Prepare Data

```{r load-data, message=FALSE, warning=FALSE}
# Load Zillow rental price data
zillow_df <- read_csv('/Users/ryoo/Documents/internship/homelessness_dataset/data/zillow_house_price.csv')

# Koreatown ZIP codes
ktown_zips <- c('90004', '90005', '90006', '90010', '90019', '90020')

# Filter for Koreatown ZIP codes
ktown_zillow <- zillow_df %>%
  filter(as.character(RegionName) %in% ktown_zips)

head(ktown_zillow)
```

```{r load-homeless-data, message=FALSE, warning=FALSE}
# Load homelessness data
homeless_df <- read_csv('data/homelessness_data/all_year.csv')

cat("Homelessness data shape:", nrow(homeless_df), "x", ncol(homeless_df), "\n")
cat("\nColumns in homelessness data:\n")
print(names(homeless_df))
cat("\nFirst few rows:\n")
print(head(homeless_df))
```


```{r zip-tract-mapping, message=FALSE, warning=FALSE}
# Get ZIP to Census Tract mapping for Koreatown using HUD API
cat("Fetching ZIP to Census Tract Mapping...\n")
url <- "https://www.huduser.gov/hudapi/public/usps?type=1&query=CA"
token <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiOWZkMDRiYjQyOTA4ZWY4ZGRlYmM1M2Y2MmRmMTU4N2I4NzdiYmNhNGFjY2ExMTgwOWJhZmEwYjZkYzFiMjUxYTRjYzhjNDc2YTU1YTg1ZjgiLCJpYXQiOjE3NjQ3MzA2MzEuNTQzMjMsIm5iZiI6MTc2NDczMDYzMS41NDMyMzMsImV4cCI6MjA4MDI2MzQzMS41MzkxNjYsInN1YiI6IjExNDI2NSIsInNjb3BlcyI6W119.AyDAhc-GSpJ8PHtURarQon645PFiPalETUQGSQmwzxxmKdwMbfc8XSj5M0tpAmsbRZEc-OZ_qI11PnjPv4Thdw"

response <- GET(url, add_headers(Authorization = paste("Bearer", token)))

if (status_code(response) == 200) {
  crosswalk_data <- content(response, "parsed")
  crosswalk_df <- as.data.frame(do.call(rbind, crosswalk_data$data$results))
  crosswalk_df <- crosswalk_df %>%
    mutate_all(unlist) %>%
    mutate(zip = as.character(zip))
  
  ktown_crosswalk <- crosswalk_df %>%
    filter(zip %in% ktown_zips) %>%
    mutate(tract = substr(as.character(geoid), nchar(as.character(geoid)) - 5, nchar(as.character(geoid))))
  
  cat("Found", nrow(ktown_crosswalk), "census tract mappings for Koreatown\n")
  cat("\nSample mappings:\n")
  print(head(ktown_crosswalk[, c('zip', 'geoid', 'tract', 'res_ratio')], 10))
} else {
  cat("Error fetching crosswalk data:", status_code(response), "\n")
  ktown_crosswalk <- NULL
}
```

```{r filter-homeless-data}
# Filter homelessness data for Koreatown census tracts
if (!is.null(ktown_crosswalk)) {
  ktown_tracts <- unique(ktown_crosswalk$tract)
  
  # Extract tract from geoid20 (last 6 digits)
  homeless_df <- homeless_df %>%
    mutate(tract = substr(as.character(geoid20), nchar(as.character(geoid20)) - 5, nchar(as.character(geoid20))))
  
  ktown_homeless <- homeless_df %>%
    filter(tract %in% ktown_tracts)
  
  # Diagnostic: Check what years we have data for
  cat("\n=== DATA DIAGNOSTIC ===\n")
  cat("Available years in homeless data:\n")
  print(sort(unique(ktown_homeless$year)))
  cat("\nCounts by year:\n")
  print(ktown_homeless %>% 
          group_by(year) %>% 
          summarise(
            n_tracts = n(),
            total_homeless = sum(tot_homeless_count, na.rm = TRUE),
            total_unsheltered = sum(unsheltered_pop_count, na.rm = TRUE)
          ))
  
  head(ktown_homeless)
}
```

## 2. Prepare Time-Series Data by COVID Period

**Time Periods:** - Before COVID: Pre-2020 (2015-2019) - During COVID:
2020-2021 - After COVID: 2022-2024

```{r aggregate-homeless, message=FALSE}
# Aggregate homelessness data by year for Koreatown
homeless_yearly <- ktown_homeless %>%
  group_by(year) %>%
  summarise(
    tot_homeless_count = sum(tot_homeless_count, na.rm = TRUE),
    unsheltered_pop_count = sum(unsheltered_pop_count, na.rm = TRUE),
    sheltered_pop_count = sum(sheltered_pop_count, na.rm = TRUE),
    livingonstreet_count = sum(livingonstreet_count, na.rm = TRUE),
    totStreetSingAdult = sum(totStreetSingAdult, na.rm = TRUE),
    totStreetFamMem = sum(totStreetFamMem, na.rm = TRUE),
    tot_vehicles_count = sum(tot_vehicles_count, na.rm = TRUE),
    tot_tent_encamp_count = sum(tot_tent_encamp_count, na.rm = TRUE),
    denom_total_pop = sum(denom_total_pop, na.rm = TRUE)
  )

# Add COVID period labels
get_covid_period <- function(year) {
  ifelse(year < 2020, 'Before COVID (2015-2019)',
         ifelse(year <= 2021, 'During COVID (2020-2021)', 'After COVID (2022+)'))
}

homeless_yearly <- homeless_yearly %>%
  mutate(period = get_covid_period(year))

cat("Homelessness data aggregated by year:\n")
print(homeless_yearly)
```

```{r process-zillow-data, message=FALSE}
# Reshape Zillow data from wide to long format
# Get date columns (those that match year-month pattern)
date_cols <- names(ktown_zillow)[grepl("^\\d{4}-\\d{2}-\\d{2}$", names(ktown_zillow))]

# Melt the data
zillow_long <- ktown_zillow %>%
  select(RegionID, RegionName, City, State, all_of(date_cols)) %>%
  pivot_longer(cols = all_of(date_cols), names_to = "date", values_to = "rent_price") %>%
  mutate(
    date = as.Date(date),
    year = year(date),
    month = month(date)
  )

# Calculate yearly average rental price for each ZIP
zillow_yearly <- zillow_long %>%
  group_by(RegionName, year) %>%
  summarise(rent_price = mean(rent_price, na.rm = TRUE), .groups = 'drop')

# Calculate overall Koreatown average by year
ktown_rent_yearly <- zillow_yearly %>%
  group_by(year) %>%
  summarise(avg_rent_price = mean(rent_price, na.rm = TRUE)) %>%
  mutate(period = get_covid_period(year))

cat("Rental price data by year:\n")
print(ktown_rent_yearly)

cat("\n=== ZILLOW DATA DIAGNOSTIC ===\n")
cat("Zillow years available:\n")
print(sort(unique(ktown_rent_yearly$year)))
```

```{r merge-data}
# Merge rental and homelessness data (aggregated yearly)
analysis_df <- homeless_yearly %>%
  inner_join(ktown_rent_yearly, by = c("year", "period"))

cat("\n=== MERGE DIAGNOSTIC ===\n")
cat("Years after merge (inner join):\n")
print(sort(unique(analysis_df$year)))
cat("\nThis shows which years have BOTH homeless AND rental data\n")
cat("If 2024 is missing here, it means either homeless or rental data is missing for 2024\n")

# Calculate rates per 100k population
analysis_df <- analysis_df %>%
  mutate(
    unsheltered_rate = (unsheltered_pop_count / denom_total_pop) * 100000,
    vehicle_rate = (tot_vehicles_count / denom_total_pop) * 100000,
    tent_rate = (tot_tent_encamp_count / denom_total_pop) * 100000,
    street_rate = (livingonstreet_count / denom_total_pop) * 100000,
    single_adult_rate = (totStreetSingAdult / denom_total_pop) * 100000,
    family_rate = (totStreetFamMem / denom_total_pop) * 100000
  )

cat("Merged analysis data:\n")
print(analysis_df %>% 
        select(year, period, avg_rent_price, unsheltered_pop_count, 
               tot_vehicles_count, tot_tent_encamp_count, livingonstreet_count))
```

```{r prepare-raw-data}
# Prepare RAW tract-level data for visualization
raw_analysis_df <- ktown_homeless %>%
  # Add period labels
  mutate(period = get_covid_period(year)) %>%
  # Merge with rental prices by year
  left_join(ktown_rent_yearly %>% select(year, avg_rent_price), by = "year") %>%
  # Calculate rates per 100k population for each tract
  mutate(
    unsheltered_rate = (unsheltered_pop_count / denom_total_pop) * 100000,
    vehicle_rate = (tot_vehicles_count / denom_total_pop) * 100000,
    tent_rate = (tot_tent_encamp_count / denom_total_pop) * 100000,
    street_rate = (livingonstreet_count / denom_total_pop) * 100000,
    single_adult_rate = (totStreetSingAdult / denom_total_pop) * 100000,
    family_rate = (totStreetFamMem / denom_total_pop) * 100000
  ) %>%
  # Remove rows with NA or infinite rates
  filter(!is.na(avg_rent_price), !is.infinite(vehicle_rate), !is.infinite(tent_rate),
         !is.infinite(street_rate), !is.infinite(unsheltered_rate))

cat("\nRaw tract-level data prepared for visualization:\n")
cat("Total observations:", nrow(raw_analysis_df), "\n")
cat("Unique census tracts:", length(unique(raw_analysis_df$tract)), "\n")
cat("Years covered:", paste(sort(unique(raw_analysis_df$year)), collapse = ", "), "\n")
```

## 3. Analysis - Question 1: Rental Costs vs. Unsheltered Homelessness Categories

Examining the relationship between rental prices and different types of
unsheltered homelessness (vehicles, tents/encampments, street) across
COVID periods.

```{r correlations-q1}
# Calculate correlations by COVID period for Question 1
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("CORRELATION ANALYSIS: Rental Costs vs. Unsheltered Homelessness Categories\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

for (period in unique(analysis_df$period)) {
  period_data <- analysis_df %>% filter(period == !!period)
  
  if (nrow(period_data) >= 2) {  # Need at least 2 points for correlation
    cat(paste0(period, ":\n"))
    cat("  Sample size:", nrow(period_data), "years\n")
    
    # Correlation with vehicle homelessness
    if (sum(period_data$tot_vehicles_count) > 0) {
      corr_vehicle <- cor(period_data$avg_rent_price, period_data$vehicle_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Vehicle homelessness: r = %.3f\n", corr_vehicle))
    }
    
    # Correlation with tent/encampment homelessness
    if (sum(period_data$tot_tent_encamp_count) > 0) {
      corr_tent <- cor(period_data$avg_rent_price, period_data$tent_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Tent/Encampment: r = %.3f\n", corr_tent))
    }
    
    # Correlation with street homelessness
    if (sum(period_data$livingonstreet_count) > 0) {
      corr_street <- cor(period_data$avg_rent_price, period_data$street_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Street: r = %.3f\n", corr_street))
    }
    cat("\n")
  }
}

cat(paste(rep("=", 80), collapse = ""), "\n")
```

```{r viz-q1-timeseries, fig.width=14, fig.height=8}
# Visualization 1: Single time series graph with all categories and dual y-axes
period_colors <- c('Before COVID (2015-2019)' = '#3498db',
                   'During COVID (2020-2021)' = '#e74c3c',
                   'After COVID (2022+)' = '#2ecc71')

# Calculate scaling factor for dual y-axis
max_rent <- max(analysis_df$avg_rent_price, na.rm = TRUE)
max_count <- max(c(analysis_df$tot_vehicles_count, analysis_df$tot_tent_encamp_count, 
                   analysis_df$livingonstreet_count, analysis_df$unsheltered_pop_count), na.rm = TRUE)

# Create combined plot
p1 <- ggplot(analysis_df, aes(x = year)) +
  # Homelessness categories (left y-axis)
  geom_line(aes(y = tot_vehicles_count, color = "Vehicle"), linewidth = 1.2) +
  geom_point(aes(y = tot_vehicles_count, color = "Vehicle"), size = 3) +
  geom_line(aes(y = tot_tent_encamp_count, color = "Tent/Encampment"), linewidth = 1.2) +
  geom_point(aes(y = tot_tent_encamp_count, color = "Tent/Encampment"), size = 3) +
  geom_line(aes(y = livingonstreet_count, color = "Street"), linewidth = 1.2) +
  geom_point(aes(y = livingonstreet_count, color = "Street"), size = 3) +
  # Rental price (right y-axis, scaled)
  geom_line(aes(y = avg_rent_price * max_count / max_rent, color = "Rental Price"), 
            linewidth = 1.5, linetype = "dashed") +
  geom_point(aes(y = avg_rent_price * max_count / max_rent, color = "Rental Price"), 
             size = 4, shape = 17) +
  scale_y_continuous(
    name = "Homelessness Count",
    sec.axis = sec_axis(~ . * max_rent / max_count, name = "Average Rental Price ($)")
  ) +
  scale_color_manual(values = c(
    "Vehicle" = "#3498db",
    "Tent/Encampment" = "#2ecc71",
    "Street" = "#9b59b6",
    "Rental Price" = "#e74c3c"
  )) +
  labs(x = "Year", 
       title = "Homelessness Categories and Rental Costs Over Time in Koreatown",
       color = "Category") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        axis.title.y.left = element_text(color = "black", face = "bold"),
        axis.title.y.right = element_text(color = "#e74c3c", face = "bold"))

print(p1)

ggsave('figures/q1_timeseries.png', width = 14, height = 8, dpi = 300)
cat("Figure saved to: figures/q1_timeseries.png\n")
```

## 4. Analysis - Question 2: Single-Adult vs. Family Homelessness

Comparing how rental costs affected single-adult versus family
homelessness rates across COVID periods.

```{r correlations-q2}
# Calculate correlations by COVID period for Question 2
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("CORRELATION ANALYSIS: Rental Costs vs. Single-Adult and Family Homelessness\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

for (period in unique(analysis_df$period)) {
  period_data <- analysis_df %>% filter(period == !!period)
  
  if (nrow(period_data) >= 2) {
    cat(paste0(period, ":\n"))
    cat("  Sample size:", nrow(period_data), "years\n")
    
    # Correlation with single adult homelessness
    if (sum(period_data$totStreetSingAdult) > 0) {
      corr_single <- cor(period_data$avg_rent_price, period_data$single_adult_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Single-Adult homelessness: r = %.3f\n", corr_single))
    }
    
    # Correlation with family homelessness
    if (sum(period_data$totStreetFamMem) > 0) {
      corr_family <- cor(period_data$avg_rent_price, period_data$family_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Family homelessness: r = %.3f\n", corr_family))
    }
    
    # Calculate ratio
    total_single <- sum(period_data$totStreetSingAdult)
    total_family <- sum(period_data$totStreetFamMem)
    cat("  Single Adults:", total_single, "\n")
    cat("  Family Members:", total_family, "\n")
    if (total_family > 0) {
      ratio <- total_single / total_family
      cat(sprintf("  Single-to-Family Ratio: %.2f\n", ratio))
    }
    cat("\n")
  }
}

cat(paste(rep("=", 80), collapse = ""), "\n")
```

```{r viz-q2-single-family}
# Visualization 3: Single Adult vs Family Homelessness Comparison
# Note: R implementation combines multiple plots similar to Python version

# Plot 1: Time series comparison (dual axis)
p1 <- ggplot(analysis_df, aes(x = year)) +
  geom_line(aes(y = avg_rent_price / 10, color = "Avg Rental Price"), linewidth = 1.5) +
  geom_point(aes(y = avg_rent_price / 10, color = "Avg Rental Price"), size = 4) +
  geom_line(aes(y = totStreetSingAdult, color = "Single Adults"), linewidth = 1.2) +
  geom_point(aes(y = totStreetSingAdult, color = "Single Adults"), size = 3, shape = 15) +
  scale_y_continuous(
    name = "Homeless Count",
    sec.axis = sec_axis(~.*10, name = "Average Rental Price ($)")
  ) +
  scale_color_manual(values = c("Avg Rental Price" = "#e74c3c",
                                 "Single Adults" = "#3498db")) +
  labs(x = "Year", title = "Rental Price vs. Single-Adult Homelessness") +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.title = element_blank(),
        legend.position = "bottom")


# Plot 2: Scatter - Rent vs Single Adult (using rate)
p2 <- ggplot(analysis_df, aes(x = avg_rent_price, y = single_adult_rate, color = period)) +
  geom_point(size = 6, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Single-Adult Homelessness Rate (per 100k)",
       title = "Rent vs. Single-Adult Homelessness Rate") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

# Plot 3: Scatter - Rent vs Family (using rate)
p3 <- ggplot(analysis_df, aes(x = avg_rent_price, y = family_rate, color = period)) +
  geom_point(size = 6, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Family Homelessness Rate (per 100k)",
       title = "Rent vs. Family Homelessness Rate") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

# Plot 4: Scatter plot - Single Adult Rate vs Family Rate
p4 <- ggplot(analysis_df, aes(x = single_adult_rate, y = family_rate, color = period)) +
  geom_point(size = 6, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgray", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Single-Adult Homelessness Rate (per 100k)", 
       y = "Family Homelessness Rate (per 100k)",
       title = "Single-Adult vs. Family Homelessness Rates") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.title = element_blank(),
        legend.position = "bottom")

print(p1)
print(p2)
print(p3)
print(p4)
```

## 5. Summary Statistics and Comprehensive Analysis

```{r summary-stats}
# Create comprehensive summary table
cat(paste(rep("=", 100), collapse = ""), "\n")
cat("COMPREHENSIVE SUMMARY: Koreatown Homelessness and Rental Costs by COVID Period\n")
cat(paste(rep("=", 100), collapse = ""), "\n\n")

summary_stats <- analysis_df %>%
  group_by(period) %>%
  summarise(
    years_min = min(year),
    years_max = max(year),
    years_count = n(),
    rent_mean = mean(avg_rent_price),
    rent_min = min(avg_rent_price),
    rent_max = max(avg_rent_price),
    vehicles_sum = sum(tot_vehicles_count),
    vehicles_mean = mean(tot_vehicles_count),
    tent_sum = sum(tot_tent_encamp_count),
    tent_mean = mean(tot_tent_encamp_count),
    street_sum = sum(livingonstreet_count),
    street_mean = mean(livingonstreet_count),
    single_sum = sum(totStreetSingAdult),
    single_mean = mean(totStreetSingAdult),
    family_sum = sum(totStreetFamMem),
    family_mean = mean(totStreetFamMem),
    unsheltered_sum = sum(unsheltered_pop_count),
    unsheltered_mean = mean(unsheltered_pop_count)
  )

cat("\nSummary Statistics by COVID Period:\n")
print(summary_stats)

# Calculate percent changes
cat("\n", paste(rep("=", 100), collapse = ""), "\n")
cat("PERCENT CHANGES ACROSS PERIODS\n")
cat(paste(rep("=", 100), collapse = ""), "\n\n")

periods <- c('Before COVID (2015-2019)', 'During COVID (2020-2021)', 'After COVID (2022+)')
metrics <- list(
  'Rental Price' = 'avg_rent_price',
  'Vehicle Homelessness' = 'tot_vehicles_count',
  'Tent/Encampment' = 'tot_tent_encamp_count',
  'Street Homelessness' = 'livingonstreet_count',
  'Single Adults' = 'totStreetSingAdult',
  'Family Members' = 'totStreetFamMem'
)

for (metric_name in names(metrics)) {
  col <- metrics[[metric_name]]
  cat(paste0("\n", metric_name, ":\n"))
  
  for (i in 1:(length(periods) - 1)) {
    before_val <- mean(analysis_df %>% filter(period == periods[i]) %>% pull(!!sym(col)))
    after_val <- mean(analysis_df %>% filter(period == periods[i+1]) %>% pull(!!sym(col)))
    
    if (before_val > 0) {
      pct_change <- ((after_val - before_val) / before_val) * 100
      cat(sprintf("  %s → %s: %+.1f%%\n", periods[i], periods[i+1], pct_change))
    } else {
      cat(sprintf("  %s → %s: N/A (no data)\n", periods[i], periods[i+1]))
    }
  }
}

cat("\n", paste(rep("=", 100), collapse = ""), "\n")
```


## 6. ZIP Code Homelessness Migration Patterns

### Which ZIP Codes Experience the Most Homelessness Growth?

This section analyzes which Koreatown ZIP codes (90004, 90005, 90006, 90010) tend to see increases or concentrations of homelessness over time.

```{r tract-zip-trends}
# Add ZIP code information to each census tract
tract_with_zip <- ktown_homeless %>%
  left_join(ktown_crosswalk %>% select(tract, zip), by = "tract") %>%
  filter(!is.na(zip))

# Aggregate by census tract, ZIP code, and year
tract_yearly_trends <- tract_with_zip %>%
  group_by(tract, zip, year) %>%
  summarise(
    tot_homeless_count = sum(tot_homeless_count, na.rm = TRUE),
    unsheltered_pop_count = sum(unsheltered_pop_count, na.rm = TRUE),
    tot_vehicles_count = sum(tot_vehicles_count, na.rm = TRUE),
    tot_tent_encamp_count = sum(tot_tent_encamp_count, na.rm = TRUE),
    livingonstreet_count = sum(livingonstreet_count, na.rm = TRUE),
    denom_total_pop = sum(denom_total_pop, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    period = get_covid_period(year),
    unsheltered_rate = (unsheltered_pop_count / denom_total_pop) * 100000
  )

# Summary of trends by ZIP code over time
zip_trends_summary <- tract_yearly_trends %>%
  group_by(zip, year, period) %>%
  summarise(
    num_tracts = n_distinct(tract),
    total_homeless = sum(tot_homeless_count, na.rm = TRUE),
    total_unsheltered = sum(unsheltered_pop_count, na.rm = TRUE),
    total_vehicles = sum(tot_vehicles_count, na.rm = TRUE),
    total_tents = sum(tot_tent_encamp_count, na.rm = TRUE),
    total_street = sum(livingonstreet_count, na.rm = TRUE),
    avg_population = mean(denom_total_pop, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    unsheltered_rate = (total_unsheltered / (num_tracts * avg_population)) * 100000
  )

cat("ZIP Code Trends Summary:\n")
print(zip_trends_summary %>% arrange(zip, year))
```

```{r zip-migration-analysis}
# Calculate average homelessness growth trajectory for each ZIP
zip_trajectory <- zip_trends_summary %>%
  group_by(zip) %>%
  arrange(year) %>%
  mutate(
    homeless_change = total_homeless - lag(total_homeless),
    unsheltered_change = total_unsheltered - lag(total_unsheltered),
    vehicles_change = total_vehicles - lag(total_vehicles),
    tents_change = total_tents - lag(total_tents)
  ) %>%
  filter(!is.na(homeless_change))

# Calculate average annual change by ZIP
avg_change_by_zip <- zip_trajectory %>%
  group_by(zip) %>%
  summarise(
    avg_homeless_change = mean(homeless_change, na.rm = TRUE),
    avg_unsheltered_change = mean(unsheltered_change, na.rm = TRUE),
    avg_vehicles_change = mean(vehicles_change, na.rm = TRUE),
    avg_tents_change = mean(tents_change, na.rm = TRUE),
    total_years = n(),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_homeless_change))

cat("\n=== Average Annual Change in Homelessness by ZIP Code ===\n")
cat("(Positive values indicate increasing homelessness)\n\n")
print(avg_change_by_zip)

# Identify destination ZIP (where homelessness is increasing most)
destination_zip <- avg_change_by_zip %>%
  filter(avg_homeless_change == max(avg_homeless_change)) %>%
  pull(zip)

cat("\n=== PRIMARY DESTINATION ZIP CODE ===\n")
cat(sprintf("ZIP Code %s shows the highest average annual increase in homelessness\n", destination_zip))
cat(sprintf("Average increase: %.1f people per year\n", 
            avg_change_by_zip %>% filter(zip == destination_zip) %>% pull(avg_homeless_change)))
```

```{r zip-concentration-analysis}
# Analyze which ZIP has highest concentration in each period
zip_concentration <- zip_trends_summary %>%
  group_by(period, zip) %>%
  summarise(
    avg_homeless = mean(total_homeless, na.rm = TRUE),
    avg_unsheltered = mean(total_unsheltered, na.rm = TRUE),
    .groups = 'drop'
  )

# Find dominant ZIP in each period
dominant_zip_by_period <- zip_concentration %>%
  group_by(period) %>%
  filter(avg_homeless == max(avg_homeless)) %>%
  select(period, zip, avg_homeless)

cat("\n=== Dominant ZIP Code by Time Period ===\n")
print(dominant_zip_by_period)

# Calculate share of total homelessness by ZIP
zip_share <- zip_trends_summary %>%
  group_by(year) %>%
  mutate(
    total_ktown_homeless = sum(total_homeless, na.rm = TRUE),
    zip_share_pct = (total_homeless / total_ktown_homeless) * 100
  ) %>%
  ungroup()

avg_share_by_zip <- zip_share %>%
  group_by(zip) %>%
  summarise(
    avg_share = mean(zip_share_pct, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_share))

cat("\n=== Average Share of Koreatown Homelessness by ZIP ===\n")
print(avg_share_by_zip)

# Calculate percent change from Before COVID to After COVID
zip_period_comparison <- zip_trends_summary %>%
  group_by(zip, period) %>%
  summarise(
    avg_homeless = mean(total_homeless, na.rm = TRUE),
    avg_unsheltered = mean(total_unsheltered, na.rm = TRUE),
    avg_vehicles = mean(total_vehicles, na.rm = TRUE),
    avg_tents = mean(total_tents, na.rm = TRUE),
    avg_street = mean(total_street, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(zip, period)

zip_change <- zip_period_comparison %>%
  pivot_wider(names_from = period, values_from = c(avg_homeless, avg_unsheltered, avg_vehicles, avg_tents, avg_street)) %>%
  mutate(
    pct_change_homeless = ((`avg_homeless_After COVID (2022+)` - `avg_homeless_Before COVID (2015-2019)`) / `avg_homeless_Before COVID (2015-2019)`) * 100,
    pct_change_unsheltered = ((`avg_unsheltered_After COVID (2022+)` - `avg_unsheltered_Before COVID (2015-2019)`) / `avg_unsheltered_Before COVID (2015-2019)`) * 100,
    pct_change_vehicles = ((`avg_vehicles_After COVID (2022+)` - `avg_vehicles_Before COVID (2015-2019)`) / `avg_vehicles_Before COVID (2015-2019)`) * 100,
    pct_change_tents = ((`avg_tents_After COVID (2022+)` - `avg_tents_Before COVID (2015-2019)`) / `avg_tents_Before COVID (2015-2019)`) * 100
  )

cat("\nPercent Change from Before COVID to After COVID by ZIP:\n")
print(zip_change %>% select(zip, starts_with("pct_change")))
```

```{r viz-growth-trajectory-enhanced}
# Enhanced Growth Trajectory Visualization
# Calculate cumulative change to show overall trajectory
zip_cumulative <- zip_trajectory %>%
  group_by(zip) %>%
  arrange(year) %>%
  mutate(
    cumulative_change = cumsum(homeless_change),
    trend_direction = ifelse(homeless_change > 0, "Growth", "Decline")
  ) %>%
  ungroup()

# Calculate overall trend for each ZIP
zip_trend_summary <- zip_trajectory %>%
  group_by(zip) %>%
  summarise(
    total_change = sum(homeless_change, na.rm = TRUE),
    avg_change = mean(homeless_change, na.rm = TRUE),
    years_growing = sum(homeless_change > 0, na.rm = TRUE),
    years_declining = sum(homeless_change < 0, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(
    net_direction = ifelse(total_change > 0, "Net Destination", "Net Source"),
    trend_label = sprintf("%s\nNet: %+.0f\nAvg: %+.1f/yr", 
                          zip, total_change, avg_change)
  )

cat("\n=== TRAJECTORY SUMMARY ===\n")
print(zip_trend_summary %>% arrange(desc(total_change)))

# Plot 2: Cumulative Change Over Time
p2_cumulative <- ggplot(zip_cumulative, aes(x = year, y = cumulative_change, color = zip, group = zip)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.8) +
  geom_line(linewidth = 2, alpha = 0.8) +
  geom_point(size = 4) +
  geom_text(data = zip_cumulative %>% filter(year == max(year)),
            aes(label = sprintf("%s: %+.0f", zip, cumulative_change)),
            hjust = -0.1, fontface = "bold", size = 3.5) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Cumulative Change: Which ZIP Codes Have Gained/Lost the Most?",
    subtitle = "Shows total net change from baseline - upward trend = destination ZIP",
    x = "Year",
    y = "Cumulative Change in Homeless Count",
    color = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

ggsave('figures/growth_trajectory_enhanced.png', width = 12, height = 8, dpi = 300)
# Plot 3: Net Change Summary Bar Chart
p3_net_change <- ggplot(zip_trend_summary, 
                        aes(x = reorder(zip, total_change), y = total_change, fill = net_direction)) +
  geom_bar(stat = "identity", alpha = 0.8, width = 0.7) +
  geom_text(aes(label = sprintf("%+.0f", total_change)), 
            hjust = ifelse(zip_trend_summary$total_change > 0, -0.2, 1.2),
            fontface = "bold", size = 5) +
  scale_fill_manual(values = c("Net Destination" = "#e74c3c", "Net Source" = "#3498db")) +
  coord_flip() +
  labs(
    title = "Net Change in Homelessness by ZIP Code (Total Period)",
    subtitle = "Positive = Destination ZIP (gaining homeless population)",
    x = "ZIP Code",
    y = "Total Net Change in Homeless Count",
    fill = "Classification"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    legend.position = "top",
    axis.text.y = element_text(size = 12, face = "bold")
  )

print(p2_cumulative)
print(p3_net_change)

cat("\nFigure saved to: figures/growth_trajectory_enhanced.png\n")

# Print interpretation
cat("\n=== INTERPRETATION ===\n")
destination_zips <- zip_trend_summary %>% filter(total_change > 0) %>% arrange(desc(total_change))
if (nrow(destination_zips) > 0) {
  cat("\nDESTINATION ZIP CODES (Net Growth):\n")
  for (i in 1:nrow(destination_zips)) {
    cat(sprintf("  %d. ZIP %s: +%.0f homeless (Avg: %+.1f/year, %d years growing)\n",
                i, 
                destination_zips$zip[i],
                destination_zips$total_change[i],
                destination_zips$avg_change[i],
                destination_zips$years_growing[i]))
  }
}

source_zips <- zip_trend_summary %>% filter(total_change < 0) %>% arrange(total_change)
if (nrow(source_zips) > 0) {
  cat("\nSOURCE ZIP CODES (Net Decline):\n")
  for (i in 1:nrow(source_zips)) {
    cat(sprintf("  %d. ZIP %s: %.0f homeless (Avg: %+.1f/year, %d years declining)\n",
                i,
                source_zips$zip[i],
                source_zips$total_change[i],
                source_zips$avg_change[i],
                source_zips$years_declining[i]))
  }
}
```

```{r viz-migration-patterns}
# Original visualization kept for reference
p_growth_trajectory <- ggplot(zip_trajectory, aes(x = year, y = homeless_change, color = zip, group = zip)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Annual Change in Homelessness by ZIP Code",
    subtitle = "Positive values indicate growth; negative values indicate decline",
    x = "Year",
    y = "Annual Change in Total Homeless Count",
    color = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "right"
  )

# Share over time
p_share_evolution <- ggplot(zip_share, aes(x = year, y = zip_share_pct, fill = zip)) +
  geom_area(alpha = 0.7) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Share of Koreatown Homelessness by ZIP Code Over Time",
    x = "Year",
    y = "Percentage of Total Koreatown Homelessness",
    fill = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

# Time series by ZIP code
p_zip_unsheltered <- ggplot(zip_trends_summary, aes(x = year, y = total_unsheltered, color = zip, group = zip)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Unsheltered Homelessness Trends by ZIP Code",
    x = "Year",
    y = "Total Unsheltered Count",
    color = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

p_zip_vehicles <- ggplot(zip_trends_summary, aes(x = year, y = total_vehicles, color = zip, group = zip)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Vehicle Homelessness Trends by ZIP Code",
    x = "Year",
    y = "Total Vehicle Count",
    color = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

# Bar chart of average change
p_avg_change <- ggplot(avg_change_by_zip, aes(x = reorder(zip, avg_homeless_change), y = avg_homeless_change, fill = zip)) +
  geom_bar(stat = "identity", alpha = 0.8) +
  geom_text(aes(label = sprintf("%.1f", avg_homeless_change)), 
            hjust = ifelse(avg_change_by_zip$avg_homeless_change > 0, -0.2, 1.2),
            fontface = "bold") +
  scale_fill_brewer(palette = "Set1") +
  coord_flip() +
  labs(
    title = "Average Annual Change in Homelessness by ZIP Code",
    x = "ZIP Code",
    y = "Average Annual Change (people/year)",
    fill = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none"
  )

# Heatmap of change by ZIP
if (nrow(zip_change) > 0) {
  p_heatmap <- ggplot(
    zip_change %>% 
      select(zip, starts_with("pct_change")) %>%
      pivot_longer(cols = starts_with("pct_change"), names_to = "metric", values_to = "pct_change") %>%
      mutate(metric = gsub("pct_change_", "", metric)),
    aes(x = metric, y = zip, fill = pct_change)
  ) +
    geom_tile(color = "white") +
    geom_text(aes(label = sprintf("%.1f%%", pct_change)), color = "white", fontface = "bold") +
    scale_fill_gradient2(low = "#2ecc71", mid = "white", high = "#e74c3c", midpoint = 0) +
    labs(
      title = "Percent Change in Homelessness: Before COVID → After COVID",
      x = "Metric",
      y = "ZIP Code",
      fill = "% Change"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
} else {
  p_heatmap <- ggplot() + labs(title = "Insufficient data for heatmap")
}

print(p_growth_trajectory)
print(p_share_evolution)
print(p_zip_unsheltered)
print(p_zip_vehicles)
```

```{r tract-ranking}
# Identify which ZIP codes had the highest increases
cat("\n=== ZIP Code Rankings by Homelessness Growth ===\n\n")

zip_ranking <- zip_change %>%
  select(zip, starts_with("pct_change")) %>%
  arrange(desc(pct_change_homeless))

cat("ZIP Codes Ranked by Total Homeless Growth:\n")
print(zip_ranking)

# Most affected census tracts
top_tracts <- tract_yearly_trends %>%
  filter(period == "After COVID (2022+)") %>%
  group_by(tract, zip) %>%
  summarise(
    avg_homeless = mean(tot_homeless_count, na.rm = TRUE),
    avg_unsheltered_rate = mean(unsheltered_rate, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  arrange(desc(avg_homeless)) %>%
  head(10)

cat("\nTop 10 Census Tracts by Homelessness (After COVID):\n")
print(top_tracts)
```

### Spatial Distribution by ZIP Code

```{r spatial-distribution-data}
# Prepare spatial distribution data for each ZIP code
# Calculate distribution metrics by ZIP and period
spatial_dist <- tract_yearly_trends %>%
  group_by(zip, period) %>%
  summarise(
    num_tracts = n_distinct(tract),
    total_homeless = sum(tot_homeless_count, na.rm = TRUE),
    total_unsheltered = sum(unsheltered_pop_count, na.rm = TRUE),
    total_vehicles = sum(tot_vehicles_count, na.rm = TRUE),
    total_tents = sum(tot_tent_encamp_count, na.rm = TRUE),
    total_street = sum(livingonstreet_count, na.rm = TRUE),
    avg_per_tract_homeless = mean(tot_homeless_count, na.rm = TRUE),
    avg_per_tract_unsheltered = mean(unsheltered_pop_count, na.rm = TRUE),
    sd_per_tract_homeless = sd(tot_homeless_count, na.rm = TRUE),
    .groups = 'drop'
  )

cat("\n=== Spatial Distribution by ZIP Code and Period ===\n")
print(spatial_dist)

# Calculate concentration index (coefficient of variation)
spatial_dist <- spatial_dist %>%
  mutate(
    concentration_index = sd_per_tract_homeless / avg_per_tract_homeless,
    concentration_category = case_when(
      is.na(concentration_index) | is.infinite(concentration_index) ~ "Insufficient Data",
      concentration_index < 0.5 ~ "Low Concentration (Evenly Distributed)",
      concentration_index < 1.0 ~ "Moderate Concentration",
      TRUE ~ "High Concentration (Clustered)"
    )
  )

cat("\n=== Concentration Analysis ===\n")
cat("Concentration Index: Higher values indicate homelessness is concentrated in fewer tracts\n\n")
print(spatial_dist %>% select(zip, period, num_tracts, concentration_index, concentration_category))
```


```{r viz-spatial-boxplots}
# Box plots showing distribution within each ZIP code
p_dist_homeless <- ggplot(tract_yearly_trends, 
                          aes(x = factor(zip), y = tot_homeless_count, fill = period)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  scale_fill_manual(values = period_colors) +
  labs(
    title = "Distribution of Homelessness Across Census Tracts by ZIP Code",
    x = "ZIP Code",
    y = "Total Homeless Count per Tract",
    fill = "Period"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

p_dist_unsheltered <- ggplot(tract_yearly_trends, 
                             aes(x = factor(zip), y = unsheltered_pop_count, fill = period)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  scale_fill_manual(values = period_colors) +
  labs(
    title = "Distribution of Unsheltered Population Across Census Tracts by ZIP Code",
    x = "ZIP Code",
    y = "Unsheltered Count per Tract",
    fill = "Period"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

p_dist_vehicles <- ggplot(tract_yearly_trends, 
                          aes(x = factor(zip), y = tot_vehicles_count, fill = period)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  scale_fill_manual(values = period_colors) +
  labs(
    title = "Distribution of Vehicle Homelessness Across Census Tracts by ZIP Code",
    x = "ZIP Code",
    y = "Vehicle Count per Tract",
    fill = "Period"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

# grid.arrange(p_dist_homeless, p_dist_unsheltered, p_dist_vehicles, ncol = 1,
#              top = grid::textGrob("Spatial Distribution: Homelessness Across Census Tracts",
#                                   gp = grid::gpar(fontsize = 16, fontface = "bold")))
print(p_dist_homeless)
print(p_dist_unsheltered)
print(p_dist_vehicles)
```

```{r viz-spatial-concentration}
# Visualization of concentration index by ZIP code
p_concentration <- ggplot(spatial_dist %>% filter(!is.na(concentration_index), !is.infinite(concentration_index)), 
                          aes(x = zip, y = concentration_index, fill = period)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "blue", linewidth = 0.8) +
  geom_hline(yintercept = 1.0, linetype = "dashed", color = "red", linewidth = 0.8) +
  scale_fill_manual(values = period_colors) +
  labs(
    title = "Concentration Index by ZIP Code and Period",
    subtitle = "Higher values = more concentrated in fewer tracts; Lower values = more evenly distributed",
    x = "ZIP Code",
    y = "Concentration Index (Coefficient of Variation)",
    fill = "Period"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom"
  ) +
  annotate("text", x = 0.5, y = 0.5, label = "Even Distribution", 
           angle = 0, vjust = -0.5, color = "blue", size = 3) +
  annotate("text", x = 0.5, y = 1.0, label = "High Concentration", 
           angle = 0, vjust = -0.5, color = "red", size = 3)

print(p_concentration)
```

```{r viz-spatial-ridgeline}
# Ridge plots showing distribution density by ZIP code (if ggridges is available)
# Install ggridges if not already installed
if (!require("ggridges", quietly = TRUE)) {
  install.packages("ggridges")
  library(ggridges)
}

p_ridge_homeless <- ggplot(tract_yearly_trends, 
                           aes(x = tot_homeless_count, y = factor(zip), fill = zip)) +
  geom_density_ridges(alpha = 0.7, scale = 2) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~ period, ncol = 1) +
  labs(
    title = "Density Distribution of Homelessness Across Census Tracts by ZIP Code",
    x = "Total Homeless Count per Tract",
    y = "ZIP Code",
    fill = "ZIP Code"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "none"
  )

print(p_ridge_homeless)
```

```{r spatial-summary-stats}
# Summary statistics for spatial distribution
cat("\n=== SPATIAL DISTRIBUTION SUMMARY ===\n\n")

for (z in unique(spatial_dist$zip)) {
  cat(paste0("\n", strrep("=", 60), "\n"))
  cat(paste0("ZIP CODE: ", z, "\n"))
  cat(paste0(strrep("=", 60), "\n\n"))
  
  zip_data <- spatial_dist %>% filter(zip == z)
  
  for (p in unique(zip_data$period)) {
    period_data <- zip_data %>% filter(period == p)
    
    cat(paste0(p, ":\n"))
    cat(sprintf("  Number of Census Tracts: %d\n", period_data$num_tracts))
    cat(sprintf("  Total Homeless: %d\n", period_data$total_homeless))
    cat(sprintf("  Average per Tract: %.1f\n", period_data$avg_per_tract_homeless))
    cat(sprintf("  Std Deviation: %.1f\n", period_data$sd_per_tract_homeless))
    cat(sprintf("  Concentration Index: %.2f (%s)\n", 
                period_data$concentration_index, 
                period_data$concentration_category))
    cat("\n")
  }
}

cat(paste0(strrep("=", 60), "\n"))
```

### Key Findings Summary

**Homelessness Destination Trends:**

1. **Primary Destination ZIP**: The analysis identifies which ZIP code experiences the highest average annual increase in homelessness, indicating where unsheltered individuals tend to concentrate over time.

2. **Period-Specific Patterns**: Different ZIP codes may dominate during different periods (Before COVID, During COVID, After COVID), revealing how displacement patterns shifted during the pandemic.

3. **Share Evolution**: By tracking each ZIP's share of total Koreatown homelessness over time, we can see whether homelessness is concentrating in specific areas or spreading more evenly.

4. **Growth Trajectories**: Year-over-year changes show which ZIP codes are experiencing sustained growth versus temporary spikes or declines in homelessness.

5. **Spatial Distribution**: The concentration index reveals whether homelessness is clustered in a few census tracts or spread more evenly across each ZIP code. Box plots and heatmaps visualize the geographic distribution patterns.

**Implications:**
- ZIP codes with consistent positive growth may need targeted interventions and resources
- Understanding destination patterns helps allocate services where they're most needed
- Identifies whether homelessness is clustering in specific neighborhoods or dispersing across Koreatown
- High concentration indices indicate specific census tracts that need focused attention
- Even distribution may require broader, ZIP-wide intervention strategies


