---
title: "Koreatown Homelessness and Rental Cost Analysis"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

# Research Questions:

1.  How are increases in rental housing costs associated with changes in
    different categories of unsheltered homelessness (vehicle,
    tent/encampment, street) in Koreatown before, during, and after
    COVID-19?
2.  Did rising rental costs affect rates of single-adult vs. family
    homelessness differently in Koreatown across the same time periods?

## Koreatown ZIP Codes: 90004, 90005, 90006, 90010

```{r setup, message=FALSE, warning=FALSE}
# Import necessary libraries
library(tidyverse)
library(httr)
library(jsonlite)
library(scales)
library(gridExtra)
library(reshape2)
library(corrplot)

# Set plotting theme
theme_set(theme_minimal())

print("Libraries imported successfully!")
```

## 1. Load and Prepare Data

```{r load-data, message=FALSE, warning=FALSE}
# Load Zillow rental price data
zillow_df <- read_csv('/Users/ryoo/Documents/internship/homelessness_dataset/data/zillow_house_price.csv')

# Koreatown ZIP codes
ktown_zips <- c('90004', '90005', '90006', '90010')

# Filter for Koreatown ZIP codes
ktown_zillow <- zillow_df %>%
  filter(as.character(RegionName) %in% ktown_zips)

head(ktown_zillow)
```

```{r load-homeless-data, message=FALSE, warning=FALSE}
# Load homelessness data
homeless_df <- read_csv('data/homelessness_data/all_year.csv')

cat("Homelessness data shape:", nrow(homeless_df), "x", ncol(homeless_df), "\n")
cat("\nColumns in homelessness data:\n")
print(names(homeless_df))
cat("\nFirst few rows:\n")
print(head(homeless_df))
```

```{r zip-tract-mapping, message=FALSE, warning=FALSE}
# Get ZIP to Census Tract mapping for Koreatown using HUD API
cat("Fetching ZIP to Census Tract Mapping...\n")
url <- "https://www.huduser.gov/hudapi/public/usps?type=1&query=CA"
token <- "eyJ0eXAiOiJKV1QiLCJhbGciOiJSUzI1NiJ9.eyJhdWQiOiI2IiwianRpIjoiOWZkMDRiYjQyOTA4ZWY4ZGRlYmM1M2Y2MmRmMTU4N2I4NzdiYmNhNGFjY2ExMTgwOWJhZmEwYjZkYzFiMjUxYTRjYzhjNDc2YTU1YTg1ZjgiLCJpYXQiOjE3NjQ3MzA2MzEuNTQzMjMsIm5iZiI6MTc2NDczMDYzMS41NDMyMzMsImV4cCI6MjA4MDI2MzQzMS41MzkxNjYsInN1YiI6IjExNDI2NSIsInNjb3BlcyI6W119.AyDAhc-GSpJ8PHtURarQon645PFiPalETUQGSQmwzxxmKdwMbfc8XSj5M0tpAmsbRZEc-OZ_qI11PnjPv4Thdw"

response <- GET(url, add_headers(Authorization = paste("Bearer", token)))

if (status_code(response) == 200) {
  crosswalk_data <- content(response, "parsed")
  crosswalk_df <- as.data.frame(do.call(rbind, crosswalk_data$data$results))
  crosswalk_df <- crosswalk_df %>%
    mutate_all(unlist) %>%
    mutate(zip = as.character(zip))
  
  ktown_crosswalk <- crosswalk_df %>%
    filter(zip %in% ktown_zips) %>%
    mutate(tract = substr(as.character(geoid), nchar(as.character(geoid)) - 5, nchar(as.character(geoid))))
  
  cat("Found", nrow(ktown_crosswalk), "census tract mappings for Koreatown\n")
  cat("\nSample mappings:\n")
  print(head(ktown_crosswalk[, c('zip', 'geoid', 'tract', 'res_ratio')], 10))
} else {
  cat("Error fetching crosswalk data:", status_code(response), "\n")
  ktown_crosswalk <- NULL
}
```

```{r filter-homeless-data}
# Filter homelessness data for Koreatown census tracts
if (!is.null(ktown_crosswalk)) {
  ktown_tracts <- unique(ktown_crosswalk$tract)
  
  # Extract tract from geoid20 (last 6 digits)
  homeless_df <- homeless_df %>%
    mutate(tract = substr(as.character(geoid20), nchar(as.character(geoid20)) - 5, nchar(as.character(geoid20))))
  
  ktown_homeless <- homeless_df %>%
    filter(tract %in% ktown_tracts)
  
  head(ktown_homeless)
}
```

## 2. Prepare Time-Series Data by COVID Period

**Time Periods:** - Before COVID: Pre-2020 (2015-2019) - During COVID:
2020-2021 - After COVID: 2022-2024

```{r aggregate-homeless, message=FALSE}
# Aggregate homelessness data by year for Koreatown
homeless_yearly <- ktown_homeless %>%
  group_by(year) %>%
  summarise(
    tot_homeless_count = sum(tot_homeless_count, na.rm = TRUE),
    unsheltered_pop_count = sum(unsheltered_pop_count, na.rm = TRUE),
    sheltered_pop_count = sum(sheltered_pop_count, na.rm = TRUE),
    livingonstreet_count = sum(livingonstreet_count, na.rm = TRUE),
    totStreetSingAdult = sum(totStreetSingAdult, na.rm = TRUE),
    totStreetFamMem = sum(totStreetFamMem, na.rm = TRUE),
    tot_vehicles_count = sum(tot_vehicles_count, na.rm = TRUE),
    tot_tent_encamp_count = sum(tot_tent_encamp_count, na.rm = TRUE),
    denom_total_pop = sum(denom_total_pop, na.rm = TRUE)
  )

# Add COVID period labels
get_covid_period <- function(year) {
  ifelse(year < 2020, 'Before COVID (2015-2019)',
         ifelse(year <= 2021, 'During COVID (2020-2021)', 'After COVID (2022+)'))
}

homeless_yearly <- homeless_yearly %>%
  mutate(period = get_covid_period(year))

cat("Homelessness data aggregated by year:\n")
print(homeless_yearly)
```

```{r process-zillow-data, message=FALSE}
# Reshape Zillow data from wide to long format
# Get date columns (those that match year-month pattern)
date_cols <- names(ktown_zillow)[grepl("^\\d{4}-\\d{2}-\\d{2}$", names(ktown_zillow))]

# Melt the data
zillow_long <- ktown_zillow %>%
  select(RegionID, RegionName, City, State, all_of(date_cols)) %>%
  pivot_longer(cols = all_of(date_cols), names_to = "date", values_to = "rent_price") %>%
  mutate(
    date = as.Date(date),
    year = year(date),
    month = month(date)
  )

# Calculate yearly average rental price for each ZIP
zillow_yearly <- zillow_long %>%
  group_by(RegionName, year) %>%
  summarise(rent_price = mean(rent_price, na.rm = TRUE), .groups = 'drop')

# Calculate overall Koreatown average by year
ktown_rent_yearly <- zillow_yearly %>%
  group_by(year) %>%
  summarise(avg_rent_price = mean(rent_price, na.rm = TRUE)) %>%
  mutate(period = get_covid_period(year))

cat("Rental price data by year:\n")
print(ktown_rent_yearly)
```

```{r merge-data}
# Merge rental and homelessness data (aggregated yearly)
analysis_df <- homeless_yearly %>%
  inner_join(ktown_rent_yearly, by = c("year", "period"))

# Calculate rates per 100k population
analysis_df <- analysis_df %>%
  mutate(
    unsheltered_rate = (unsheltered_pop_count / denom_total_pop) * 100000,
    vehicle_rate = (tot_vehicles_count / denom_total_pop) * 100000,
    tent_rate = (tot_tent_encamp_count / denom_total_pop) * 100000,
    street_rate = (livingonstreet_count / denom_total_pop) * 100000,
    single_adult_rate = (totStreetSingAdult / denom_total_pop) * 100000,
    family_rate = (totStreetFamMem / denom_total_pop) * 100000
  )

cat("Merged analysis data:\n")
print(analysis_df %>% 
        select(year, period, avg_rent_price, unsheltered_pop_count, 
               tot_vehicles_count, tot_tent_encamp_count, livingonstreet_count))
```

```{r prepare-raw-data}
# Prepare RAW tract-level data for visualization
raw_analysis_df <- ktown_homeless %>%
  # Add period labels
  mutate(period = get_covid_period(year)) %>%
  # Merge with rental prices by year
  left_join(ktown_rent_yearly %>% select(year, avg_rent_price), by = "year") %>%
  # Calculate rates per 100k population for each tract
  mutate(
    unsheltered_rate = (unsheltered_pop_count / denom_total_pop) * 100000,
    vehicle_rate = (tot_vehicles_count / denom_total_pop) * 100000,
    tent_rate = (tot_tent_encamp_count / denom_total_pop) * 100000,
    street_rate = (livingonstreet_count / denom_total_pop) * 100000,
    single_adult_rate = (totStreetSingAdult / denom_total_pop) * 100000,
    family_rate = (totStreetFamMem / denom_total_pop) * 100000
  ) %>%
  # Remove rows with NA or infinite rates
  filter(!is.na(avg_rent_price), !is.infinite(vehicle_rate), !is.infinite(tent_rate),
         !is.infinite(street_rate), !is.infinite(unsheltered_rate))

cat("\nRaw tract-level data prepared for visualization:\n")
cat("Total observations:", nrow(raw_analysis_df), "\n")
cat("Unique census tracts:", length(unique(raw_analysis_df$tract)), "\n")
cat("Years covered:", paste(sort(unique(raw_analysis_df$year)), collapse = ", "), "\n")
```

## 3. Analysis - Question 1: Rental Costs vs. Unsheltered Homelessness Categories

Examining the relationship between rental prices and different types of
unsheltered homelessness (vehicles, tents/encampments, street) across
COVID periods.

```{r correlations-q1}
# Calculate correlations by COVID period for Question 1
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("CORRELATION ANALYSIS: Rental Costs vs. Unsheltered Homelessness Categories\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

for (period in unique(analysis_df$period)) {
  period_data <- analysis_df %>% filter(period == !!period)
  
  if (nrow(period_data) >= 2) {  # Need at least 2 points for correlation
    cat(paste0(period, ":\n"))
    cat("  Sample size:", nrow(period_data), "years\n")
    
    # Correlation with vehicle homelessness
    if (sum(period_data$tot_vehicles_count) > 0) {
      corr_vehicle <- cor(period_data$avg_rent_price, period_data$vehicle_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Vehicle homelessness: r = %.3f\n", corr_vehicle))
    }
    
    # Correlation with tent/encampment homelessness
    if (sum(period_data$tot_tent_encamp_count) > 0) {
      corr_tent <- cor(period_data$avg_rent_price, period_data$tent_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Tent/Encampment: r = %.3f\n", corr_tent))
    }
    
    # Correlation with street homelessness
    if (sum(period_data$livingonstreet_count) > 0) {
      corr_street <- cor(period_data$avg_rent_price, period_data$street_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Street: r = %.3f\n", corr_street))
    }
    cat("\n")
  }
}

cat(paste(rep("=", 80), collapse = ""), "\n")
```

```{r viz-q1-timeseries, fig.width=14, fig.height=8}
# Visualization 1: Single time series graph with all categories and dual y-axes
period_colors <- c('Before COVID (2015-2019)' = '#3498db',
                   'During COVID (2020-2021)' = '#e74c3c',
                   'After COVID (2022+)' = '#2ecc71')

# Calculate scaling factor for dual y-axis
max_rent <- max(analysis_df$avg_rent_price, na.rm = TRUE)
max_count <- max(c(analysis_df$tot_vehicles_count, analysis_df$tot_tent_encamp_count, 
                   analysis_df$livingonstreet_count, analysis_df$unsheltered_pop_count), na.rm = TRUE)

# Create combined plot
p1 <- ggplot(analysis_df, aes(x = year)) +
  # Homelessness categories (left y-axis)
  geom_line(aes(y = tot_vehicles_count, color = "Vehicle"), linewidth = 1.2) +
  geom_point(aes(y = tot_vehicles_count, color = "Vehicle"), size = 3) +
  geom_line(aes(y = tot_tent_encamp_count, color = "Tent/Encampment"), linewidth = 1.2) +
  geom_point(aes(y = tot_tent_encamp_count, color = "Tent/Encampment"), size = 3) +
  geom_line(aes(y = livingonstreet_count, color = "Street"), linewidth = 1.2) +
  geom_point(aes(y = livingonstreet_count, color = "Street"), size = 3) +
  # Rental price (right y-axis, scaled)
  geom_line(aes(y = avg_rent_price * max_count / max_rent, color = "Rental Price"), 
            linewidth = 1.5, linetype = "dashed") +
  geom_point(aes(y = avg_rent_price * max_count / max_rent, color = "Rental Price"), 
             size = 4, shape = 17) +
  scale_y_continuous(
    name = "Homelessness Count",
    sec.axis = sec_axis(~ . * max_rent / max_count, name = "Average Rental Price ($)")
  ) +
  scale_color_manual(values = c(
    "Vehicle" = "#3498db",
    "Tent/Encampment" = "#2ecc71",
    "Street" = "#9b59b6",
    "Rental Price" = "#e74c3c"
  )) +
  labs(x = "Year", 
       title = "Homelessness Categories and Rental Costs Over Time in Koreatown",
       color = "Category") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        axis.title.y.left = element_text(color = "black", face = "bold"),
        axis.title.y.right = element_text(color = "#e74c3c", face = "bold"))

print(p1)

ggsave('figures/q1_timeseries.png', width = 14, height = 8, dpi = 300)
cat("Figure saved to: figures/q1_timeseries.png\n")
```

```{r viz-q1-correlations, fig.width=18, fig.height=5}
# Visualization 2: Scatter plots showing correlation between rent and each category (using rates)
p1 <- ggplot(analysis_df, aes(x = avg_rent_price, y = vehicle_rate, color = period)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Vehicle Homelessness Rate (per 100k)",
       title = "Rent vs. Vehicle Homelessness Rate") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

p2 <- ggplot(analysis_df, aes(x = avg_rent_price, y = tent_rate, color = period)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Tent/Encampment Rate (per 100k)",
       title = "Rent vs. Tent/Encampment Rate") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

p3 <- ggplot(analysis_df, aes(x = avg_rent_price, y = street_rate, color = period)) +
  geom_point(size = 5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Street Homelessness Count",
       title = "Rent vs. Street Homelessness") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

grid.arrange(p1, p2, p3, ncol = 3,
             top = grid::textGrob("Correlation: Rental Prices vs. Unsheltered Homelessness Types",
                                  gp = grid::gpar(fontsize = 16, fontface = "bold")))

ggsave('figures/q1_correlations.png', width = 18, height = 5, dpi = 300)
cat("Figure saved to: figures/q1_correlations.png\n")
```

```{r viz-percentage-change, fig.width=14, fig.height=8}
# Visualization: Percentage Change from Baseline (2015)
# Calculate percentage change from the first year (baseline)
baseline_year <- min(analysis_df$year)
baseline_data <- analysis_df %>% filter(year == baseline_year)

pct_change_df <- analysis_df %>%
  mutate(
    pct_rent = ((avg_rent_price - baseline_data$avg_rent_price[1]) / baseline_data$avg_rent_price[1]) * 100,
    pct_vehicle = ((tot_vehicles_count - baseline_data$tot_vehicles_count[1]) / 
                    ifelse(baseline_data$tot_vehicles_count[1] == 0, 1, baseline_data$tot_vehicles_count[1])) * 100,
    pct_tent = ((tot_tent_encamp_count - baseline_data$tot_tent_encamp_count[1]) / 
                 ifelse(baseline_data$tot_tent_encamp_count[1] == 0, 1, baseline_data$tot_tent_encamp_count[1])) * 100,
    pct_street = ((livingonstreet_count - baseline_data$livingonstreet_count[1]) / 
                   ifelse(baseline_data$livingonstreet_count[1] == 0, 1, baseline_data$livingonstreet_count[1])) * 100,
    pct_unsheltered = ((unsheltered_pop_count - baseline_data$unsheltered_pop_count[1]) / 
                        baseline_data$unsheltered_pop_count[1]) * 100
  )

# Reshape for plotting
pct_long <- pct_change_df %>%
  select(year, pct_rent, pct_vehicle, pct_tent, pct_street) %>%
  pivot_longer(cols = -year, names_to = "category", values_to = "pct_change") %>%
  mutate(category = recode(category,
                           "pct_rent" = "Rental Price",
                           "pct_vehicle" = "Vehicle",
                           "pct_tent" = "Tent/Encampment",
                           "pct_street" = "Street",
                           "pct_unsheltered" = "Total Unsheltered"))

# Create percentage change plot
ggplot(pct_long, aes(x = year, y = pct_change, color = category, group = category)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c(
    "Rental Price" = "#e74c3c",
    "Vehicle" = "#3498db",
    "Tent/Encampment" = "#2ecc71",
    "Street" = "#9b59b6",
    "Total Unsheltered" = "#f39c12"
  )) +
  labs(x = "Year", 
       y = paste0("Percentage Change from Baseline (", baseline_year, ")"),
       title = paste0("Percentage Change in Rental Prices and Homelessness Categories (Baseline: ", baseline_year, ")"),
       color = "Category") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        legend.position = "bottom",
        legend.title = element_text(face = "bold"),
        panel.grid.major = element_line(color = "gray90"),
        panel.grid.minor = element_line(color = "gray95")) +
  scale_y_continuous(labels = function(x) paste0(x, "%"))

ggsave('figures/percentage_change.png', width = 14, height = 8, dpi = 300)
cat("Figure saved to: figures/percentage_change.png\n")
```

## 4. Analysis - Question 2: Single-Adult vs. Family Homelessness

Comparing how rental costs affected single-adult versus family
homelessness rates across COVID periods.

```{r correlations-q2}
# Calculate correlations by COVID period for Question 2
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("CORRELATION ANALYSIS: Rental Costs vs. Single-Adult and Family Homelessness\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

for (period in unique(analysis_df$period)) {
  period_data <- analysis_df %>% filter(period == !!period)
  
  if (nrow(period_data) >= 2) {
    cat(paste0(period, ":\n"))
    cat("  Sample size:", nrow(period_data), "years\n")
    
    # Correlation with single adult homelessness
    if (sum(period_data$totStreetSingAdult) > 0) {
      corr_single <- cor(period_data$avg_rent_price, period_data$single_adult_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Single-Adult homelessness: r = %.3f\n", corr_single))
    }
    
    # Correlation with family homelessness
    if (sum(period_data$totStreetFamMem) > 0) {
      corr_family <- cor(period_data$avg_rent_price, period_data$family_rate, use = "complete.obs")
      cat(sprintf("  Rent vs Family homelessness: r = %.3f\n", corr_family))
    }
    
    # Calculate ratio
    total_single <- sum(period_data$totStreetSingAdult)
    total_family <- sum(period_data$totStreetFamMem)
    cat("  Single Adults:", total_single, "\n")
    cat("  Family Members:", total_family, "\n")
    if (total_family > 0) {
      ratio <- total_single / total_family
      cat(sprintf("  Single-to-Family Ratio: %.2f\n", ratio))
    }
    cat("\n")
  }
}

cat(paste(rep("=", 80), collapse = ""), "\n")
```

```{r viz-q2-single-family, fig.width=16, fig.height=12}
# Visualization 3: Single Adult vs Family Homelessness Comparison
# Note: R implementation combines multiple plots similar to Python version

# Plot 1: Time series comparison (dual axis)
p1 <- ggplot(analysis_df, aes(x = year)) +
  geom_line(aes(y = avg_rent_price / 10, color = "Avg Rental Price"), linewidth = 1.5) +
  geom_point(aes(y = avg_rent_price / 10, color = "Avg Rental Price"), size = 4) +
  geom_line(aes(y = totStreetSingAdult, color = "Single Adults"), linewidth = 1.2) +
  geom_point(aes(y = totStreetSingAdult, color = "Single Adults"), size = 3, shape = 15) +
  geom_line(aes(y = totStreetFamMem, color = "Family"), linewidth = 1.2) +
  geom_point(aes(y = totStreetFamMem, color = "Family"), size = 3, shape = 17) +
  scale_y_continuous(
    name = "Homeless Count",
    sec.axis = sec_axis(~.*10, name = "Average Rental Price ($)")
  ) +
  scale_color_manual(values = c("Avg Rental Price" = "#e74c3c",
                                 "Single Adults" = "#3498db",
                                 "Family" = "#2ecc71")) +
  labs(x = "Year", title = "Rental Price vs. Single-Adult & Family Homelessness") +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.title = element_blank(),
        legend.position = "bottom")

ggsave('figures/q2_single_vs_family1.png')

# Plot 2: Scatter - Rent vs Single Adult (using rate)
p2 <- ggplot(analysis_df, aes(x = avg_rent_price, y = single_adult_rate, color = period)) +
  geom_point(size = 6, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Single-Adult Homelessness Rate (per 100k)",
       title = "Rent vs. Single-Adult Homelessness Rate") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

# Plot 3: Scatter - Rent vs Family (using rate)
p3 <- ggplot(analysis_df, aes(x = avg_rent_price, y = family_rate, color = period)) +
  geom_point(size = 6, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Average Rental Price ($)", y = "Family Homelessness Rate (per 100k)",
       title = "Rent vs. Family Homelessness Rate") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.position = "bottom")

# Plot 4: Scatter plot - Single Adult Rate vs Family Rate
p4 <- ggplot(analysis_df, aes(x = single_adult_rate, y = family_rate, color = period)) +
  geom_point(size = 6, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgray", linetype = "dashed", alpha = 0.5) +
  scale_color_manual(values = period_colors) +
  labs(x = "Single-Adult Homelessness Rate (per 100k)", 
       y = "Family Homelessness Rate (per 100k)",
       title = "Single-Adult vs. Family Homelessness Rates") +
  theme(plot.title = element_text(face = "bold", size = 13),
        legend.title = element_blank(),
        legend.position = "bottom")

grid.arrange(p1, p2, p3, p4, ncol = 2,
             top = grid::textGrob("Single-Adult vs. Family Homelessness and Rental Costs in Koreatown",
                                  gp = grid::gpar(fontsize = 16, fontface = "bold")))

cat("Figure saved to: figures/q2_single_vs_family.png\n")
```

## 5. Summary Statistics and Comprehensive Analysis

```{r summary-stats}
# Create comprehensive summary table
cat(paste(rep("=", 100), collapse = ""), "\n")
cat("COMPREHENSIVE SUMMARY: Koreatown Homelessness and Rental Costs by COVID Period\n")
cat(paste(rep("=", 100), collapse = ""), "\n\n")

summary_stats <- analysis_df %>%
  group_by(period) %>%
  summarise(
    years_min = min(year),
    years_max = max(year),
    years_count = n(),
    rent_mean = mean(avg_rent_price),
    rent_min = min(avg_rent_price),
    rent_max = max(avg_rent_price),
    vehicles_sum = sum(tot_vehicles_count),
    vehicles_mean = mean(tot_vehicles_count),
    tent_sum = sum(tot_tent_encamp_count),
    tent_mean = mean(tot_tent_encamp_count),
    street_sum = sum(livingonstreet_count),
    street_mean = mean(livingonstreet_count),
    single_sum = sum(totStreetSingAdult),
    single_mean = mean(totStreetSingAdult),
    family_sum = sum(totStreetFamMem),
    family_mean = mean(totStreetFamMem),
    unsheltered_sum = sum(unsheltered_pop_count),
    unsheltered_mean = mean(unsheltered_pop_count)
  )

cat("\nSummary Statistics by COVID Period:\n")
print(summary_stats)

# Calculate percent changes
cat("\n", paste(rep("=", 100), collapse = ""), "\n")
cat("PERCENT CHANGES ACROSS PERIODS\n")
cat(paste(rep("=", 100), collapse = ""), "\n\n")

periods <- c('Before COVID (2015-2019)', 'During COVID (2020-2021)', 'After COVID (2022+)')
metrics <- list(
  'Rental Price' = 'avg_rent_price',
  'Vehicle Homelessness' = 'tot_vehicles_count',
  'Tent/Encampment' = 'tot_tent_encamp_count',
  'Street Homelessness' = 'livingonstreet_count',
  'Single Adults' = 'totStreetSingAdult',
  'Family Members' = 'totStreetFamMem'
)

for (metric_name in names(metrics)) {
  col <- metrics[[metric_name]]
  cat(paste0("\n", metric_name, ":\n"))
  
  for (i in 1:(length(periods) - 1)) {
    before_val <- mean(analysis_df %>% filter(period == periods[i]) %>% pull(!!sym(col)))
    after_val <- mean(analysis_df %>% filter(period == periods[i+1]) %>% pull(!!sym(col)))
    
    if (before_val > 0) {
      pct_change <- ((after_val - before_val) / before_val) * 100
      cat(sprintf("  %s → %s: %+.1f%%\n", periods[i], periods[i+1], pct_change))
    } else {
      cat(sprintf("  %s → %s: N/A (no data)\n", periods[i], periods[i+1]))
    }
  }
}

cat("\n", paste(rep("=", 100), collapse = ""), "\n")
```

```{r viz-heatmap, fig.width=12, fig.height=8}
# Create comprehensive heatmap showing all correlations
correlation_data <- data.frame()

for (period in unique(analysis_df$period)) {
  period_data <- analysis_df %>% filter(period == !!period)
  
  if (nrow(period_data) >= 2) {
    corr_vehicle <- cor(period_data$avg_rent_price, period_data$vehicle_rate, use = "complete.obs")
    corr_tent <- cor(period_data$avg_rent_price, period_data$tent_rate, use = "complete.obs")
    corr_street <- cor(period_data$avg_rent_price, period_data$street_rate, use = "complete.obs")
    corr_single <- cor(period_data$avg_rent_price, period_data$single_adult_rate, use = "complete.obs")
    corr_family <- ifelse(sum(period_data$totStreetFamMem) > 0,
                          cor(period_data$avg_rent_price, period_data$family_rate, use = "complete.obs"),
                          NA)
    
    correlation_data <- rbind(correlation_data, data.frame(
      period = period,
      Vehicle = corr_vehicle,
      `Tent/Encampment` = corr_tent,
      Street = corr_street,
      `Single Adult` = corr_single,
      Family = corr_family,
      check.names = FALSE
    ))
  }
}

# Create matrix for heatmap
corr_matrix <- as.matrix(correlation_data[, -1])
rownames(corr_matrix) <- gsub(" ", "\n", correlation_data$period)

# Melt for ggplot
corr_melted <- melt(corr_matrix)
names(corr_melted) <- c("Period", "Category", "Correlation")

ggplot(corr_melted, aes(x = Category, y = Period, fill = Correlation)) +
  geom_tile(color = "white", linewidth = 2) +
  geom_text(aes(label = sprintf("%.3f", Correlation)), size = 5) +
  scale_fill_gradient2(low = "#d73027", mid = "#ffffbf", high = "#1a9850",
                       midpoint = 0, limit = c(-1, 1),
                       name = "Correlation\nCoefficient") +
  labs(title = "Correlation: Rental Prices vs. Homelessness Rates (per 100k) by COVID Period",
       x = "Homelessness Category",
       y = "COVID Period") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
        axis.title = element_text(face = "bold", size = 13),
        axis.text.x = element_text(size = 11),
        axis.text.y = element_text(size = 11))

ggsave('figures/correlation_heatmap.png', width = 12, height = 8, dpi = 300)
cat("Figure saved to: figures/correlation_heatmap.png\n")
```

```{r viz-comprehensive, fig.width=18, fig.height=10, warning=FALSE}
# Create a combined visualization showing rental price changes and homelessness trends
# Note: This is a simplified version of the Python comprehensive plot
# R's grid system can be used for more complex layouts if needed

# Main dual-axis plot
p_main <- ggplot(analysis_df, aes(x = year)) +
  geom_ribbon(aes(ymin = 0, ymax = avg_rent_price / 50), fill = "#e74c3c", alpha = 0.2) +
  geom_line(aes(y = avg_rent_price / 50, color = "Rental Price"), linewidth = 2) +
  geom_point(aes(y = avg_rent_price / 50, color = "Rental Price"), size = 5) +
  geom_ribbon(aes(ymin = 0, ymax = unsheltered_pop_count), fill = "#3498db", alpha = 0.2) +
  geom_line(aes(y = unsheltered_pop_count, color = "Unsheltered"), linewidth = 2) +
  geom_point(aes(y = unsheltered_pop_count, color = "Unsheltered"), size = 5, shape = 15) +
  scale_color_manual(values = c("Rental Price" = "#e74c3c", "Unsheltered" = "#3498db")) +
  scale_y_continuous(
    name = "Unsheltered Homelessness Count",
    sec.axis = sec_axis(~.*50, name = "Average Rental Price ($)")
  ) +
  labs(x = "Year",
       title = "Koreatown: Rental Prices and Unsheltered Homelessness Trends") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16),
        legend.title = element_blank(),
        legend.position = "top")

# Category breakdowns
p_vehicle <- ggplot(analysis_df, aes(x = factor(year), y = tot_vehicles_count, fill = period)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = period_colors) +
  labs(x = "Year", y = "Count", title = "Vehicle Homelessness") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave('figures/vehicle.png')

p_tent <- ggplot(analysis_df, aes(x = factor(year), y = tot_tent_encamp_count, fill = period)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = period_colors) +
  labs(x = "Year", y = "Count", title = "Tent/Encampment") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave('figures/tent.png')

p_street <- ggplot(analysis_df, aes(x = factor(year), y = livingonstreet_count, fill = period)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = period_colors) +
  labs(x = "Year", y = "Count", title = "Street Homelessness") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave('figures/street.png')

# Single vs Family trend
p_single_family <- ggplot(analysis_df, aes(x = year)) +
  geom_line(aes(y = totStreetSingAdult, color = "Single Adults"), linewidth = 1.5) +
  geom_point(aes(y = totStreetSingAdult, color = "Single Adults"), size = 4) +
  geom_line(aes(y = totStreetFamMem, color = "Family Members"), linewidth = 1.5) +
  geom_point(aes(y = totStreetFamMem, color = "Family Members"), size = 4, shape = 15) +
  scale_color_manual(values = c("Single Adults" = "#9b59b6", "Family Members" = "#f39c12")) +
  labs(x = "Year", y = "Count", title = "Single-Adult vs. Family Homelessness") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.title = element_blank(),
        legend.position = "bottom")

# Arrange all plots
layout_matrix <- rbind(c(1, 1, 1),
                       c(2, 3, 4),
                       c(5, 5, 5))

grid.arrange(p_main, p_vehicle, p_tent, p_street, p_single_family,
             layout_matrix = layout_matrix,
             top = grid::textGrob("Comprehensive Analysis: Koreatown Homelessness and Rental Costs",
                                  gp = grid::gpar(fontsize = 18, fontface = "bold")))

ggsave('figures/comprehensive_analysis.png', width = 18, height = 10, dpi = 300)
cat("Figure saved to: figures/comprehensive_analysis.png\n")
```

## 6. Key Findings and Conclusions

### Research Question 1: Rental Costs vs. Unsheltered Homelessness Categories

**Key Findings:** - The analysis examines the relationship between
rental housing costs and three categories of unsheltered homelessness
(vehicle, tent/encampment, street) across three COVID periods -
Correlation coefficients and trend visualizations reveal how different
types of unsheltered homelessness respond to rental price changes -
Time-series graphs show the evolution of each category before, during,
and after COVID-19

### Research Question 2: Single-Adult vs. Family Homelessness

**Key Findings:** - Comparison of how rental costs affected single-adult
versus family homelessness differently across COVID periods - Ratio
analysis shows the relative prevalence of single adults compared to
families - Correlation analysis reveals whether single adults or
families were more vulnerable to rental price increases

### Overall Insights:

-   Visual evidence from scatter plots and time series shows trends
    across different periods
-   The heatmap provides a comprehensive view of all correlations at a
    glance
-   Period-specific analysis reveals how the pandemic may have altered
    the relationship between housing costs and homelessness
